# Using Fonts in R: A Comprehensive Guide (Claude 3.7)

## Table of Contents

<h3>
1. [Introduction](#introduction)
2. [Font Basics in R](#font-basics-in-r)
3. [Using System Fonts](#using-system-fonts)
4. [Adding Custom Fonts](#adding-custom-fonts)
5. [Fonts in Different R Graphics Systems](#fonts-in-different-r-graphics-systems)
   - [Base R](#base-r)
   - [ggplot2](#ggplot2)
   - [lattice](#lattice)
6. [Font Manipulation with showtext](#font-manipulation-with-showtext)
7. [Working with Non-Latin Characters](#working-with-non-latin-characters)
8. [Troubleshooting Font Issues](#troubleshooting-font-issues)
9. [Examples](#examples)
10. [Resources](#resources)
</h2>

## Introduction

<h3>
Fonts play a crucial role in data visualization, affecting both readability and visual appeal. R offers several ways to customize fonts in your plots, graphs, and other visualizations. This guide covers how to use, manage, and customize fonts in R across different graphics systems.
</h3>

## Font Basics in R

<h3>
In R, fonts are handled differently depending on the graphics device and package you're using. The three main approaches to graphics in R each have their own way of handling fonts:

- **Base R**: Uses device-dependent font settings
- **ggplot2**: Uses themes for font customization
- **lattice**: Uses parameter settings

Default font families typically available in R include:

- "serif" (similar to Times New Roman)
- "sans" (similar to Arial or Helvetica)
- "mono" (similar to Courier)
</h3>

## Using System Fonts

### Checking Available Fonts

<h3>
To see which fonts are available on your system:

```{r}
# With extrafont package
library(extrafont)
fonts()

# On Windows
#windowsFonts()

# On Mac
#quartzFonts()

# On Unix/Linux
#X11Fonts()
```

</h3>

### Loading System Fonts

<h3>
The extrafont package helps R locate and use system fonts:

```{r}
# Install if needed
# install.packages("extrafont")

library(extrafont)
# Import all system fonts (only need to do once)
font_import()
# Load fonts for the R session
loadfonts()
```

</h3>

## Adding Custom Fonts

### Google Fonts with `showtext`

```{r}
# Install if needed
# install.packages("showtext")

library(showtext)

# Add Google fonts
font_add_google("Roboto", "roboto")
font_add_google("Montserrat", "montserrat")

# Enable showtext
showtext_auto()
```

### Local Font Files

```r
library(showtext)

# Add a font from a file
font_add(family = "my_custom_font", 
         regular = "path/to/font.ttf",
         bold = "path/to/font-bold.ttf",
         italic = "path/to/font-italic.ttf",
         bolditalic = "path/to/font-bolditalic.ttf")

# Enable showtext
showtext_auto()
```

## Fonts in Different R Graphics Systems

### Base R

```{r}
# Set font family for a plot
plot(1:10, main = "Title", family = "serif")

# Customize text with a different font
text(5, 5, "Text Label", family = "sans")

# Using Windows-specific fonts
if (.Platform$OS.type == "windows") {
  windowsFonts(myFont = windowsFont("Arial"))
  plot(1:10, main = "Title", family = "myFont")
}
```

### ggplot2

```{r}
library(ggplot2)

# Basic font family setting
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "MPG vs Weight") +
  theme(text = element_text(family = "serif"))

# Detailed font customization
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "MPG vs Weight", subtitle = "From mtcars dataset") +
  theme(
    plot.title = element_text(family = "roboto", face = "bold", size = 16),
    plot.subtitle = element_text(family = "roboto", face = "italic", size = 12),
    axis.title = element_text(family = "sans", size = 12),
    axis.text = element_text(family = "mono", size = 10)
  )
```

### lattice

```{r}
library(lattice)

# Set default font
trellis.par.set(fontsize = list(text = 12), 
                fontfamily = list(text = "serif"))

# Create plot with custom fonts
xyplot(mpg ~ wt, data = mtcars,
       main = list(label = "MPG vs Weight", cex = 1.5, fontfamily = "sans"),
       xlab = list(label = "Weight", cex = 1.2, fontfamily = "mono"),
       ylab = list(label = "MPG", cex = 1.2, fontfamily = "serif"))
```

## Font Manipulation with showtext

<h3>
The `showtext` package offers powerful font handling for R graphics:

```{r}
library(showtext)
library(ggplot2)

# Add and enable fonts
font_add_google("Pacifico", "pacifico")
showtext_auto()

# Create a plot with the custom font
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "MPG vs Weight",
    x = "Weight (1000 lbs)",
    y = "Miles Per Gallon"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(family = "pacifico", size = 22, hjust = 0.5),
    axis.title = element_text(family = "pacifico", size = 14),
    axis.text = element_text(family = "sans", size = 12)
  )

# Save the plot with fonts embedded
ggsave("custom_font_plot.pdf", width = 8, height = 6)
```

</h3>

## Working with Non-Latin Characters

<h3>

For languages using non-Latin scripts (Chinese, Japanese, Korean, Arabic, etc.):

```{r}
library(showtext)

# Add a font with CJK support
font_add_google("Noto Sans SC", "noto-sc")  # For Simplified Chinese
showtext_auto()

# Create a plot with Chinese text
plot(1:10, main = "这是中文标题", family = "noto-sc")
text(5, 5, "中文文本", family = "noto-sc")
```

</h3>

## Troubleshooting Font Issues

<h3>
Common font issues and solutions:

1. **Fonts not showing correctly in saved plots**
   - Use `ggsave()` with `device = cairo_pdf` or `device = cairo_ps`
   - Use `showtext` package to embed fonts

2. **Missing fonts errors**
   - Check if the font is available: `fonts()`
   - Import system fonts: `font_import()`
   - Load fonts for the session: `loadfonts()`

3. **PDF export problems**
   - Use `embed_fonts()` from extrafont: 
     ```r
     ggsave("plot.pdf", plot)
     embed_fonts("plot.pdf")
   ```

</h3>

## Examples

### Combining Multiple Fonts in ggplot2

```{r}
library(ggplot2)
library(showtext)

# Add fonts
font_add_google("Oswald", "oswald")
font_add_google("Lato", "lato")
showtext_auto()

# Create plot with multiple fonts
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line(color = "#1565C0", size = 1) +
  labs(
    title = "US Unemployment",
    subtitle = "From economics dataset",
    x = "Year",
    y = "Unemployment (thousands)",
    caption = "Source: stats.bls.gov"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(family = "oswald", size = 20, face = "bold"),
    plot.subtitle = element_text(family = "lato", size = 12, face = "italic"),
    axis.title = element_text(family = "lato", size = 12),
    axis.text = element_text(family = "lato", size = 10),
    plot.caption = element_text(family = "lato", size = 8)
  )
```

### Creating a Font Comparison Plot

```{r}
library(ggplot2)
library(showtext)
library(grid)
library(gridExtra)

# Add several fonts
font_add_google("Roboto", "roboto")
font_add_google("Playfair Display", "playfair")
font_add_google("Source Code Pro", "source")
font_add_google("Open Sans", "opensans")
showtext_auto()

# Create a function to generate sample text
font_sample <- function(family, size = 12) {
  ggplot() + 
    annotate("text", x = 0, y = 0, 
             label = paste0("Aa Bb Cc 123\nThe quick brown fox: ", family),
             family = family, size = size, hjust = 0) +
    xlim(-0.1, 1) + ylim(-0.5, 0.5) +
    theme_void() +
    labs(title = family) +
    theme(plot.title = element_text(family = family, size = 14))
}

# Create sample plots for each font
p1 <- font_sample("roboto")
p2 <- font_sample("playfair")
p3 <- font_sample("source")
p4 <- font_sample("opensans")

# Combine them into a single comparison plot
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Resources

### Useful Packages

<h3>
- **extrafont**: Import and use system fonts
- **showtext**: Add and render custom fonts including Google Fonts
- **ragg**: High-quality graphics device with excellent font support
- **systemfonts**: Access system fonts

</h3>

### Further Reading
- [R Graphics Cookbook](https://r-graphics.org/) by Winston Chang
- [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/) by Hadley Wickham
- [R Graph Gallery](https://www.r-graph-gallery.com/) for visualization examples
- [Google Fonts](https://fonts.google.com/) for browsing available fonts

---

### This guide covers the basics of working with fonts in R. Remember that font availability may differ across operating systems, and consider embedding fonts when sharing plots for consistent appearance across devices.