---
output:
  html_document:
    code_folding: hide
---


# Using Fonts in R (from Grok3)

This guide explains how to work with fonts in R, focusing on base R graphics, `ggplot2`, and external font integration. Fonts can enhance visualizations by improving readability and aesthetics.

## 1. Fonts in Base R Graphics

Base R plotting functions (e.g., 'plot()', 'text()') allow font customization via the 'family' argument or graphical parameters ('par()').

### Basic Usage
- The 'family' argument specifies the font family.
- Common options: "serif", "sans", "mono".
- Example:

```{r}
plot(1:10, main = "Example Plot", family = "sans")
text(2, 8, "Hello, R!", family = "serif", cex = 1.5)
text(5, 2, "Hello, R!", family = "Helvetica", cex = 1.5)
text(8, 5, "Hello, R!", family = "ArialMT", cex = 1.5)
```

### Setting Fonts Globally
Use `par()` to set the default font for a plotting session:
```{r}
par(family = "mono")
plot(1:10, main = "Mono Font Plot")
```

### Available Fonts
To see available font families in base R:
```{r}
names(postscriptFonts())
```
This lists PostScript font mappings, though availability depends on your system.

## 2. Fonts in ggplot2

<h3>
The `ggplot2` package offers more flexibility for font customization through themes and the `extrafont` or `showtext` packages.
</h3>

### Basic Font Changes
#### Use `theme()` to adjust fonts for titles, labels, etc.:
```{r}
library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Car Weight vs MPG") +
  theme(plot.title = element_text(family = "sans", size = 16))
```

### Using System Fonts with `extrafont`
1. Install and load `extrafont`:
```{r}
   #install.packages("extrafont")
   library(extrafont)
```
2. Import system fonts (run once):
```{r}
   font_import()
```
3. List available fonts:
```{r}
   fonts()
```
4. Use a specific font in `ggplot2`:
```{r}
   ggplot(mtcars, aes(x = wt, y = mpg)) +
     geom_point() +
     labs(title = "Custom Font") +
     theme(plot.title = element_text(family = "Comic Sans MS",size=24))
```

### Using `showtext` for More Flexibility
The `showtext` package simplifies font usage, especially for online fonts (e.g., Google Fonts):
1. Install and load:
```{r}
   #install.packages("showtext")
   library(showtext)
```
2. Enable `showtext`:
```{r}
   showtext_auto()
```
3. Load a Google Font:
```{r}
   sysfonts::font_add_google("Roboto", "roboto")
```
4. Use it in a plot:
```{r}
   ggplot(mtcars, aes(x = wt, y = mpg)) +
     geom_point() +
     labs(title = "Roboto Font") +
     theme(plot.title = element_text(family = "roboto",size=24))
```

## 3. Platform-Specific Notes

- **Windows**: Fonts installed on your system are typically available.
- **MacOS**: Use `quartzFonts()` to map fonts in base R or rely on `extrafont`/`showtext`.
- **Linux**: Ensure fonts are installed via your package manager (e.g., `ttf` packages), then import them.

## 4. Tips
- Check font availability before plotting to avoid errors.
- Use `extrafont::fonttable()` or `sysfonts::font_files()` to debug font issues.
- For publication-quality plots, embed fonts when saving (e.g., with `extrafont::embed_fonts()` for PDFs).

## Example: Combining Everything
```{r}
library(ggplot2)
library(showtext)
showtext_auto()
sysfonts::font_add_google("Lato", "lato")

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Lato Font Example", x = "Weight", y = "Miles per Gallon") +
  theme(text = element_text(family = "lato",size=24))
```

```{r}
library(ggplot2)
library(showtext)
showtext_auto()
sysfonts::font_add_google("Lato", "lato")
sysfonts::font_add_google("Sigmar One","sigmar")

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Lato Font Example", x = "Weight", y = "Miles per Gallon") +
  theme(plot.title = element_text(family = "lato",size=24),
        axis.title.x = element_text(family="sigmar",size=16)
        )
```

## Adding Fonts
There are websites with many fonts, such as [google fonts](https://fonts.google.com/)

Get a list of installed Google Fonts
```{r}
font_families_google()
```

See if the font of interest is included:
```{r}
font_families_google()[grep("Roboto",font_families_google())]
```
If it is not installed
```{r}
font_families_google()[grep("Playwrite",font_families_google())]
```
Then it will need to be installed
```{r}
font_add("playwrite",regular="Playwrite_IT_Moderna/PlaywriteITModerna-VariableFont_wght.ttf")
```

And:
```{r}
font_add_google("Roboto Condensed","Roboto")
font_add_google("Noto Sans","Noto")
font_add_google("Life Savers","LS")
mtcars |>
  ggplot(aes(x=cyl,y=mpg))+
  geom_col()+
  geom_text(aes(x=6,y=200,label="label goes here"),family="Roboto")+
  labs(title="title goes here",
       x="cylinders",
       y="miles per gallon")+
  theme_bw()+
  theme(plot.title=element_text(family="LS",hjust=.5,size=24),
        axis.title.x=element_text(face="bold",family="playwrite",size=16),
        axis.title.y=element_text(family="Noto",size=14))
```

This Markdown file provides a starting point for font usage in R. Experiment with different families and packages to suit your needs!

---

